---
title: "Triple Crown Motocross - 2025 Report"
format: 
  html:
    toc: true
---

# 1 Abstract
Give an overall summary of the report. What is different from last year and what to expect.

# 2 Data scraping and tidying

```{r}
#| warning: false
#| output: false
library(tidyverse)
library(rvest)
library(stringr)
library(lubridate)
library(gt)
library(ggplot2)
library(expm)
library(purrr)
```

Here is the points structure for the series.
```{r}
points <- tibble(finish = 1:40, 
                 points = c(25,22,20,18,16,15,14,13,12,11,10,9,8,7,6,5,4,3,2,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))

points |>
  slice_head(n = 20) |>
  gt() |>
  cols_label(finish = "Finish",
             points = "Points")
```

Here is a function that converts the lap times (string format) to seconds (numeric format). For example, a lap time of 1:13.21 gets converted to 73.21.

```{r}
convert_to_seconds <- function(time_string) {
  if (is.na(time_string) | time_string == "") {
    return(NA)
  }
  
  if (grepl(":", time_string)) {
    mins <- as.numeric(substr(time_string, 1, 
                              regexpr(":", time_string) - 1))
    secs <- as.numeric(substr(time_string, 
                              regexpr(":", time_string) + 1,
                              nchar(time_string)))
    
    return(mins * 60 + secs)
  } 
  
  else {
    numeric_val <- as.numeric(time_string)
    
    return(numeric_val)
  }
}
```

Here is the function to scrap the data from the Trackside tables.
```{r}
mx_scraping <- function(url, year, round, moto, track,
                        date, raceid, class, type) {
  webpage <- read_html(url)
  lap_times_table <- webpage |>
    html_nodes("table") |>
    html_table(fill = TRUE)
  
  lap_times <- lap_times_table[[8]]
  
  lap_times <- lap_times |>
    as_tibble() |>
    slice(3:n())
  
  lap <- paste0("L", 1:(ncol(lap_times)-1))
  lap2 <- append("name", lap)
  lap_times_reduced <- lap_times |>
    setNames(lap2)
  
  check <- lap_times_reduced$name[1]
  if (str_detect(check, "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t")) {
    lap_times_reduced <- lap_times_reduced |>
      separate_wider_delim(
        name,
        delim = "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t",
        names = c("name1", "name2"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name1,
        delim = " - ",
        names = c("finish", "rider"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name2,
        delim = "\t\r\n\t\t\t\t\t ",
        names = c("location", "number"),
        too_few = "align_start"
      ) |>
      drop_na(rider) |>
      mutate(
        finish = seq_len(nrow(lap_times_reduced)-1),
        number = str_remove(number, "#")
      )
  } else {
    lap_times_reduced <- lap_times_reduced |>
      separate_wider_delim(
        name,
        delim = " - ",
        names = c("finish", "name"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name,
        delim = "#",
        names = c("rider", "number")
      ) |>
      drop_na(rider) |>
      mutate(rider = str_trim(rider),
             finish = seq_len(nrow(lap_times_reduced)-1),
             location = NA
      )
  }
  
  lap_times_reduced2 <- lap_times_reduced
  for (col in lap) {
    lap_times_reduced2 <- lap_times_reduced2 |>
      separate_wider_delim(
        col,
        delim = "\r\n\t\t\t\t\t",
        names = c(paste0(col, "_time"), paste0(col, "_behind"), 
                  paste0(col, "_place")),
        too_few = "align_start"
      )
  }
  
  lap3 <- paste0("L", 1:length(lap), "_place")
  lap_times_reduced2 <- lap_times_reduced2 |>
    mutate(across(lap3, ~ str_extract(.x, "\\d+") |> 
                    as.numeric(), .names = "{col}"))
  
  lap4 <- paste0("L", 1:length(lap), "_time")
  lap5 <- paste0("L", 1:length(lap), "_behind")
  final_lap_times <- lap_times_reduced2
  for (i in lap4) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }
  for (i in lap5) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }
  
  final_lap_times <- final_lap_times |>
    pivot_longer(
      cols = starts_with("L"), 
      names_to = c("lap", ".value"), 
      names_pattern = "L(\\d+)_(time|behind|place)",
      names_repair = "unique"
    ) |>
    mutate(lap = as.numeric(lap),
           year = year,
           round = round,
           moto = moto,
           track = track,
           date = parse_date(date, "%m/%d/%Y"),
           race_id = raceid,
           class = class,
           type = type) |>
    left_join(points) |>
    drop_na(time)
  
  return(final_lap_times)
  
}
```

```{r}
#| warning: false
rd11_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "450_2025_1", "450", "mx")

rd12_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "450_2025_2", "450", "mx")

rd21_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20773&rn=1&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/08/2025", "450_2025_3", "450", "mx")

rd22_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20773&rn=2&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "450_2025_4", "450", "mx")

rd31_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "450_2025_5", "450", "mx")

rd32_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "450_2025_6", "450", "mx")

rd41_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "450_2025_7", "450", "mx")

rd42_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "450_2025_8", "450", "mx")

rd51_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "450_2025_9", "450", "mx")

rd52_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "450_2025_10", "450", "mx")

rd61_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "450_2025_11", "450", "mx")

rd62_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "450_2025_12", "450", "mx")

rd71_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "450_2025_13", "450", "mx")

rd72_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "450_2025_14", "450", "mx")

rd81_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "450_2025_15", "450", "mx")

rd82_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "450_2025_16", "450", "mx")


all_2025_450 <- bind_rows(rd11_2025_450, rd12_2025_450,
                          rd21_2025_450, rd22_2025_450,
                          rd31_2025_450, rd32_2025_450,
                          rd41_2025_450, rd42_2025_450,
                          rd51_2025_450, rd52_2025_450,
                          rd61_2025_450, rd62_2025_450,
                          rd71_2025_450, rd72_2025_450,
                          rd81_2025_450, rd82_2025_450)

rd11_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "250_2025_1", "250", "mx")

rd12_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "250_2025_2", "250", "mx")

rd21_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20773&rn=1&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/08/2025", "250_2025_3", "250", "mx")

rd22_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20773&rn=2&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "250_2025_4", "250", "mx")

rd31_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "250_2025_5", "250", "mx")

rd32_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "250_2025_6", "250", "mx")

rd41_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "250_2025_7", "250", "mx")

rd42_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "250_2025_8", "250", "mx")

rd51_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "250_2025_9", "250", "mx")

rd52_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "250_2025_10", "250", "mx")

rd61_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "450_2025_11", "450", "mx")

rd62_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "450_2025_12", "450", "mx")

rd71_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "450_2025_13", "450", "mx")

rd72_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "450_2025_14", "450", "mx")

rd81_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "450_2025_15", "450", "mx")

rd82_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "450_2025_16", "450", "mx")

all_2025_250 <- bind_rows(rd11_2025_250, rd12_2025_250,
                          rd21_2025_250, rd22_2025_250,
                          rd31_2025_250, rd32_2025_250,
                          rd41_2025_250, rd42_2025_250,
                          rd51_2025_250, rd52_2025_250,
                          rd61_2025_250, rd62_2025_250,
                          rd71_2025_250, rd72_2025_250,
                          rd81_2025_250, rd82_2025_250)

rd11_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "wmx_2025_1", "WMX", "mx")

rd12_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "wmx_2025_2", "WMX", "mx")

rd21_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=21641&rn=2&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/07/2025", "wmx_2025_3", "WMX", "mx")

rd22_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20773&rn=1&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "wmx_2025_4", "WMX", "mx")

rd23_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20773&rn=2&rt=M", 2025, 2, 3, "Mason Watt Raceway", "06/08/2025", "wmx_2025_5", "WMX", "mx")

rd31_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "wmx_2025_6", "WMX", "mx")

rd32_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "wmx_2025_7", "WMX", "mx")

rd41_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "wmx_2025_8", "WMX", "mx")

rd42_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "wmx_2025_9", "WMX", "mx")

rd51_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "wmx_2025_10", "WMX", "mx")

rd52_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "wmx_2025_11", "WMX", "mx")

rd61_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "450_2025_11", "450", "mx")

rd62_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "450_2025_12", "450", "mx")

rd71_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "450_2025_13", "450", "mx")

rd72_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "450_2025_14", "450", "mx")

rd81_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "450_2025_15", "450", "mx")

rd82_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "450_2025_16", "450", "mx")

all_2025_wmx <- bind_rows(rd11_2025_wmx, rd12_2025_wmx,
                          rd21_2025_wmx, rd22_2025_wmx, rd23_2025_wmx,
                          rd31_2025_wmx, rd32_2025_wmx,
                          rd41_2025_wmx, rd42_2025_wmx,
                          rd51_2025_wmx, rd52_2025_wmx,
                          rd61_2025_wmx, rd62_2025_wmx,
                          rd71_2025_wmx, rd72_2025_wmx,
                          rd81_2025_wmx, rd82_2025_wmx)

all_2025 <- bind_rows(all_2025_450, all_2025_250, all_2025_wmx)
write.csv(all_2025,"all_2025.csv")
```

```{r}
#| warning: false
#| eval: false
all_2025_riders <- all_2025 |>
  distinct(rider)

# List of all riders from the Trackside website
# Type "%" in the search bar
url <- "https://cmrc.tracksideresults.com/search.asp?search=%25&searchType=name&Submit=Find"
webpage <- read_html(url)
riders_table <- webpage |>
  html_nodes("table") |>
  html_table(fill = TRUE)

riders <- riders_table[[6]]
  
riders <- riders |>
  as_tibble() |>
  slice(2:n()) |>
  rename(name = X2, number = X3, location = X4) |>
  select(name, number, location)

write.csv(riders,"all_2025_riders.csv", row.names = FALSE)
```

```{r}
#| warning: false
# Replace " " with " " and remove X column (row number)
# Save as all_2025_cleaned.csv
# In all_2025_riders.csv, added manufacturer, updated_location, round variables
# Save as all_2025_riders_complete.csv

all_2025_cleaned <- read.csv("all_2025_cleaned.csv")
all_2025_riders_complete <- read.csv("all_2025_riders_complete.csv")
```

```{r}
#| warning: false
all_2025_riders_complete <- all_2025_riders_complete |>
  select(-number)

all_2025_complete <- all_2025_cleaned |>
  cross_join(all_2025_riders_complete) |>
  filter(str_detect(rider, name)) |>
  filter(is.na(round.y) | round.x == round.y) |>
  select(finish, name, number, lap, time, behind, place, year, round.x, moto, track, date, race_id, class, type, points, manufacturer, updated_location) |>
  rename(rider = name, round = round.x, location = updated_location) |>
  distinct() |>
  mutate(date = ymd(date))

write.csv(all_2025_complete,"all_2025_complete.csv")
```

New to scraping this year are the bike manufacturers of each rider. 

In future iterations, we want to look at using AI scraping software to help with the process. The process of collecting manufacturer info for ~300 riders is tedious.

# 3 Analysis

## 3.1 Reapplying 2024 analysis

### 3.1.1 Demographic summary

Distributions of home locations:

```{r}
#| warning: false
#| echo: false
all_2025_complete |>
  distinct(class, rider, location) |>
  group_by(class, location) |>
  summarize(n = n(),
            .groups = "drop") |>
  ungroup() |>
  arrange(desc(n)) |>
  pivot_wider(names_from = class,
              values_from = n) |>
  filter(! location == "unknown") |>
  mutate(`250` = if_else(is.na(`250`), 0, `250`),
         WMX = if_else(is.na(WMX), 0, WMX),
         `450` = if_else(is.na(`450`), 0, `450`)) |>
  select(location, `450`, `250`, WMX) |>
  gt() |>
  cols_label(location = "Location",
             `450` = "450",
             `250` = "250",
             WMX = "WMX")
```

### 3.1.2 Passing probabilities

<u>450 class</u>:
```{r}
#| warning: false
#| echo: false
all_2025_450_add <- all_2025_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_450_add2 <- all_2025_450_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_450_add <- all_2025_450_add |>
  left_join(all_2025_450_add2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrix450_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_450_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_450_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix450_1[i, j] <- matrix_value
  }
}

rownames(matrix450_1) <- cats
colnames(matrix450_1) <- cats
print(round(matrix450_1, 2))
```

```{r}
#| warning: false
#| echo: false
all_2025_450_placecat <- all_2025_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, 
                            "21-25","26-30"))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", "26-30")))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_450_placecat2 <- all_2025_450_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_450_placecat <- all_2025_450_placecat |>
  left_join(all_2025_450_placecat2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrix450_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_450_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_450_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix450_2[i, j] <- matrix_value
  }
}

rownames(matrix450_2) <- cats2
colnames(matrix450_2) <- cats2

tm_450 <- as.data.frame(as.table(matrix450_2))
ggplot(tm_450, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - 450 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

<u>250 class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_250_add <- all_2025_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_250_add2 <- all_2025_250_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_250_add <- all_2025_250_add |>
  left_join(all_2025_250_add2) |>
  mutate(prob = n/total)

rows = 40
cols = 40
cats <- 1:40

matrix250_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_250_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_250_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix250_1[i, j] <- matrix_value
  }
}

rownames(matrix250_1) <- cats
colnames(matrix250_1) <- cats
print(round(matrix250_1, 2))
```

```{r}
#| warning: false
#| echo: false
all_2025_250_placecat <- all_2025_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30",
                     ifelse(place >= 31 & place <= 35,
                            "31-35", "36-40"))))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                          ifelse(place_next >= 26 & place_next <= 30, "26-30",
                          ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_250_placecat2 <- all_2025_250_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_250_placecat <- all_2025_250_placecat |>
  left_join(all_2025_250_placecat2) |>
  mutate(prob = n/total)

rows2 = 8
cols2 = 8
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
           "31-35", "36-40")

matrix250_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_250_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_250_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix250_2[i, j] <- matrix_value
  }
}

rownames(matrix250_2) <- cats2
colnames(matrix250_2) <- cats2

tm_250 <- as.data.frame(as.table(matrix250_2))
ggplot(tm_250, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - 250 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

<u>WMX class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_wmx_add <- all_2025_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_wmx_add2 <- all_2025_wmx_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_wmx_add <- all_2025_wmx_add |>
  left_join(all_2025_wmx_add2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrixwmx_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_wmx_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_wmx_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrixwmx_1[i, j] <- matrix_value
  }
}

rownames(matrixwmx_1) <- cats
colnames(matrixwmx_1) <- cats
print(round(matrixwmx_1, 2))
```

```{r}
#| warning: false
all_2025_wmx_placecat <- all_2025_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20", "21-25")))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20", "21-25"))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_wmx_placecat2 <- all_2025_wmx_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_wmx_placecat <- all_2025_wmx_placecat |>
  left_join(all_2025_wmx_placecat2) |>
  mutate(prob = n/total)

rows2 = 5
cols2 = 5
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25")

matrixwmx_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_wmx_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_wmx_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrixwmx_2[i, j] <- matrix_value
  }
}

rownames(matrixwmx_2) <- cats2
colnames(matrixwmx_2) <- cats2

tm_wmx <- as.data.frame(as.table(matrixwmx_2))
ggplot(tm_wmx, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - WMX class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

At a track level:
``` {r}
#| warning: false
#| echo: false
pass_prob_moto_fn <- function(input_class, input_track) {
  placecat <- all_2025_complete |>
    filter(class == input_class & track == input_track) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35,
                                                                        "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
    rows = distinct_cat
    cols = distinct_cat
    if (distinct_cat == 4) {
      cats <- c("1-5", "6-10", "11-15", "16-20")
    } else if (distinct_cat == 5) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
    } else if (distinct_cat == 6) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
    } else if (distinct_cat == 7) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35")
    } else if (distinct_cat == 8) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35", "36-40")
    }
    
    pass_matrix <- matrix(0, nrow = rows, ncol = cols)
    for (i in 1:rows) {
      for (j in 1:cols) {
        matrix_row <- placecat |>
          filter(place_cat == cats[i] & place_next_cat == cats[j])
        if (nrow(matrix_row) == 0) {
          matrix_value <- 0
        } else {
          matrix_value <- placecat |> 
            filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
        }
        pass_matrix[i, j] <- matrix_value
      }
    }
    
    rownames(pass_matrix) <- cats
    colnames(pass_matrix) <- cats

  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  p <- ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32", guide = "none") +
    geom_text(aes(label = round(Freq,2)), size = 4) +
    labs(title = "Passing Probabilities",
         subtitle = input_track,
         x = "From Place", 
         y = "To Place", fill = "Probability")
  
  return(p)
}

tracks1 <- all_2025_complete |>
  filter(round <= 4) |>
  distinct(track) |>
  pull(track)

tracks2 <- all_2025_complete |>
  filter(round > 4) |>
  distinct(track) |>
  pull(track)

plots1 <- map(
  tracks1,
  ~ pass_prob_moto_fn(
      input_class = "450",
      input_track = .x
    )
)

plots2 <- map(
  tracks2,
  ~ pass_prob_moto_fn(
      input_class = "450",
      input_track = .x
    )
)

library(patchwork)
wrap_plots(plots1, ncol = 2)
wrap_plots(plots2, ncol = 2)
```

At an individual rider level (Nicoletti, Ward, Pettis):
```{r}
#| warning: false
#| echo: false
pass_prob_rider_fn <- function(input_rider) {
  placecat <- all_2025_complete |>
    filter(rider == input_rider) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35, "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
  rows = distinct_cat
  cols = distinct_cat
  if (distinct_cat == 2) {
    cats <- c("1-5", "6-10")
  } else if (distinct_cat == 3) {
    cats <- c("1-5", "6-10", "11-15")
  } else if (distinct_cat == 4) {
    cats <- c("1-5", "6-10", "11-15", "16-20")
  } else if (distinct_cat == 5) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
  } else if (distinct_cat == 6) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
  } else if (distinct_cat == 7) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35")
  } else if (distinct_cat == 8) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35", "36-40")
  }
  
  pass_matrix <- matrix(0, nrow = rows, ncol = cols)
  for (i in 1:rows) {
    for (j in 1:cols) {
      matrix_row <- placecat |>
        filter(place_cat == cats[i] & place_next_cat == cats[j])
      if (nrow(matrix_row) == 0) {
        matrix_value <- 0
      } else {
        matrix_value <- placecat |> 
          filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
      }
      pass_matrix[i, j] <- matrix_value
    }
  }
  
  rownames(pass_matrix) <- cats
  colnames(pass_matrix) <- cats
  
  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  p <- ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32", guide = "none") +
    geom_text(aes(label = round(Freq,2)), size = 3) +
    labs(title = "Passing Probabilities",
         subtitle = input_rider, 
         x = "From Place", 
         y = "To Place", fill = "Probability")
  
  return(p)
}

pp_riders <- c("PHIL NICOLETTI", "TANNER WARD", "JESS PETTIS")
plots1 <- map(
  pp_riders,
  ~ pass_prob_rider_fn(
      input_rider = .x
    )
)
wrap_plots(plots1, ncol = 3)

test <- all_2025_complete |>
  filter(rider == "JESS PETTIS") |>
  distinct(place)

pass_prob_rider_fn("TANNER WARD")
```


For fatique effects, compare the first moto with the second moto of the day. Expect slower lap times. Consider mechanical changes or improved/worse track conditions.
Look at the big findings from last year - WMX passing probs and fatigue effects - and see if same trends this year.
Compare everything from last year to this year. Where is it headed? Were there improvements?
## 3.2 Moto quartile positions
Looking at a rider's position at the 0/25/50/75/100% point of the moto.
## 3.3 Standarized lap times
Need a way to compare across riders and tracks.
## 3.4 Consistency scores
Separate steady from volatile riders. Need to apply lap time standardization
## 3.5 Defensive skill
Ability to hold position under pressure (rider consistently within 2 seconds back).
## 3.6 Inflection laps
Revisiting win probability model from last year
More evaluation to make sure it is accurate
Identify laps where win probability shifts sharply
## 3.7 Traffic effects
Clear air vs battles: Top riders standardized times, split between when in 1st place and trailing someone by 2 seconds.
What about when leaders come up on lap traffic? Blue flag - does it work or are leaders slowed down?
## 3.8 Gender gap
Compare 250 class and WMX class riders since they use the same bike. How would the women do in the men's series (simulation)?
## 3.9 Prize money
Is it a financially viable occupation? Need to reference official series docs.
## 3.10 Entertainment value
Average gap to leader by series
## 3.11 Weather effects
Compare the lap times for this year to last year at the same tracks with the temperature. Consider weather types (like rain) and track changes (not the same setup every year)
## 3.12 Fall prediction
First, creating the reasonable definition for a fall - big delta in lap time and place, big difference between last lap place and finish position. No easy way to confirm.
Then look at the characteristics/conditions around those falls - which riders typically involved, late in the moto?
What would happen in the points race if certain accidents (falls of 2 riders at same time) didn't happen?
## 3.13 Rider profiles
Conducted in Tableau this time. Include link here.
Archetype clusters -> Look at the NHL/NBA player cards (percentiles) and try to replicate.
Include photos to display.
## 3.14 Expected finish model
Linear regression/random forest?
Finish ~ Rider historical pace + lap + current place + time gaps to each place
## 3.15 Pass probability model
P(place-1) ~ Rider historical pace? + lap + current place + time behind next place + time ahead of next place

# 4 Extensions
The big thing I want to do to expand the project in the future is add previous years data to the analysis. Consolidate everything in one dataset. According to the [results](https://triplecrownseries.ca/results/) tab of the Triple Crown Series website, there is data going back to 2018.

# 5 Notes