---
title: "Triple Crown Motocross - 2025 Report"
format: 
  html:
    toc: true
---

# 1 Abstract
Give an overall summary of the report. What is different from last year and what to expect.

# 2 Data scraping and tidying

```{r}
#| warning: false
#| output: false
library(tidyverse)
library(rvest)
library(stringr)
library(lubridate)
library(gt)
library(ggplot2)
library(expm)
library(purrr)
```

Here is the points structure for the series.
```{r}
points <- tibble(finish = 1:40, 
                 points = c(25,22,20,18,16,15,14,13,12,11,10,9,8,7,6,5,4,3,2,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))

points |>
  slice_head(n = 20) |>
  gt() |>
  cols_label(finish = "Finish",
             points = "Points")
```

Here is a function that converts the lap times (string format) to seconds (numeric format). For example, a lap time of 1:13.21 gets converted to 73.21.

```{r}
convert_to_seconds <- function(time_string) {
  if (is.na(time_string) | time_string == "") {
    return(NA)
  }
  
  if (grepl(":", time_string)) {
    mins <- as.numeric(substr(time_string, 1, 
                              regexpr(":", time_string) - 1))
    secs <- as.numeric(substr(time_string, 
                              regexpr(":", time_string) + 1,
                              nchar(time_string)))
    
    return(mins * 60 + secs)
  } 
  
  else {
    numeric_val <- as.numeric(time_string)
    
    return(numeric_val)
  }
}
```

Here is the function to scrap the data from the Trackside tables.
```{r}
mx_scraping <- function(url, year, round, moto, track,
                        date, raceid, class, type) {
  webpage <- read_html(url)
  lap_times_table <- webpage |>
    html_nodes("table") |>
    html_table(fill = TRUE)
  
  lap_times <- lap_times_table[[8]]
  
  lap_times <- lap_times |>
    as_tibble() |>
    slice(3:n())
  
  lap <- paste0("L", 1:(ncol(lap_times)-1))
  lap2 <- append("name", lap)
  lap_times_reduced <- lap_times |>
    setNames(lap2)
  
  check <- lap_times_reduced$name[1]
  if (str_detect(check, "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t")) {
    lap_times_reduced <- lap_times_reduced |>
      separate_wider_delim(
        name,
        delim = "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t",
        names = c("name1", "name2"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name1,
        delim = " - ",
        names = c("finish", "rider"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name2,
        delim = "\t\r\n\t\t\t\t\t ",
        names = c("location", "number"),
        too_few = "align_start"
      ) |>
      drop_na(rider) |>
      mutate(
        finish = seq_len(nrow(lap_times_reduced)-1),
        number = str_remove(number, "#")
      )
  } else {
    lap_times_reduced <- lap_times_reduced |>
      separate_wider_delim(
        name,
        delim = " - ",
        names = c("finish", "name"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name,
        delim = "#",
        names = c("rider", "number")
      ) |>
      drop_na(rider) |>
      mutate(rider = str_trim(rider),
             finish = seq_len(nrow(lap_times_reduced)-1),
             location = NA
      )
  }
  
  lap_times_reduced2 <- lap_times_reduced
  for (col in lap) {
    lap_times_reduced2 <- lap_times_reduced2 |>
      separate_wider_delim(
        col,
        delim = "\r\n\t\t\t\t\t",
        names = c(paste0(col, "_time"), paste0(col, "_behind"), 
                  paste0(col, "_place")),
        too_few = "align_start"
      )
  }
  
  lap3 <- paste0("L", 1:length(lap), "_place")
  lap_times_reduced2 <- lap_times_reduced2 |>
    mutate(across(lap3, ~ str_extract(.x, "\\d+") |> 
                    as.numeric(), .names = "{col}"))
  
  lap4 <- paste0("L", 1:length(lap), "_time")
  lap5 <- paste0("L", 1:length(lap), "_behind")
  final_lap_times <- lap_times_reduced2
  for (i in lap4) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }
  for (i in lap5) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }
  
  final_lap_times <- final_lap_times |>
    pivot_longer(
      cols = starts_with("L"), 
      names_to = c("lap", ".value"), 
      names_pattern = "L(\\d+)_(time|behind|place)",
      names_repair = "unique"
    ) |>
    mutate(lap = as.numeric(lap),
           year = year,
           round = round,
           moto = moto,
           track = track,
           date = parse_date(date, "%m/%d/%Y"),
           race_id = raceid,
           class = class,
           type = type) |>
    left_join(points) |>
    drop_na(time)
  
  return(final_lap_times)
  
}
```

```{r}
#| warning: false
rd11_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "450_2025_1", "450", "mx")

rd12_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "450_2025_2", "450", "mx")

rd21_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20773&rn=1&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/08/2025", "450_2025_3", "450", "mx")

rd22_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20773&rn=2&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "450_2025_4", "450", "mx")

rd31_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "450_2025_5", "450", "mx")

rd32_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "450_2025_6", "450", "mx")

rd41_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "450_2025_7", "450", "mx")

rd42_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "450_2025_8", "450", "mx")

rd51_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "450_2025_9", "450", "mx")

rd52_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "450_2025_10", "450", "mx")

rd61_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "450_2025_11", "450", "mx")

rd62_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "450_2025_12", "450", "mx")

rd71_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "450_2025_13", "450", "mx")

rd72_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "450_2025_14", "450", "mx")

rd81_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "450_2025_15", "450", "mx")

rd82_2025_450 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=33&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "450_2025_16", "450", "mx")


all_2025_450 <- bind_rows(rd11_2025_450, rd12_2025_450,
                          rd21_2025_450, rd22_2025_450,
                          rd31_2025_450, rd32_2025_450,
                          rd41_2025_450, rd42_2025_450,
                          rd51_2025_450, rd52_2025_450,
                          rd61_2025_450, rd62_2025_450,
                          rd71_2025_450, rd72_2025_450,
                          rd81_2025_450, rd82_2025_450)

rd11_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "250_2025_1", "250", "mx")

rd12_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "250_2025_2", "250", "mx")

rd21_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20773&rn=1&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/08/2025", "250_2025_3", "250", "mx")

rd22_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20773&rn=2&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "250_2025_4", "250", "mx")

rd31_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "250_2025_5", "250", "mx")

rd32_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "250_2025_6", "250", "mx")

rd41_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "250_2025_7", "250", "mx")

rd42_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "250_2025_8", "250", "mx")

rd51_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "250_2025_9", "250", "mx")

rd52_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "250_2025_10", "250", "mx")

rd61_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "250_2025_11", "250", "mx")

rd62_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "250_2025_12", "250", "mx")

rd71_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "250_2025_13", "250", "mx")

rd72_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "250_2025_14", "250", "mx")

rd81_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "250_2025_15", "250", "mx")

rd82_2025_250 <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=32&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "250_2025_16", "250", "mx")

all_2025_250 <- bind_rows(rd11_2025_250, rd12_2025_250,
                          rd21_2025_250, rd22_2025_250,
                          rd31_2025_250, rd32_2025_250,
                          rd41_2025_250, rd42_2025_250,
                          rd51_2025_250, rd52_2025_250,
                          rd61_2025_250, rd62_2025_250,
                          rd71_2025_250, rd72_2025_250,
                          rd81_2025_250, rd82_2025_250)

rd11_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20766&rn=1&rt=M", 2025, 1, 1, "Wild Rose MX", "06/01/2025", "wmx_2025_1", "WMX", "mx")

rd12_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20766&rn=2&rt=M", 2025, 1, 2, "Wild Rose MX", "06/01/2025", "wmx_2025_2", "WMX", "mx")

rd21_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=21641&rn=2&rt=M", 2025, 2, 1, "Mason Watt Raceway", "06/07/2025", "wmx_2025_3", "WMX", "mx")

rd22_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20773&rn=1&rt=M", 2025, 2, 2, "Mason Watt Raceway", "06/08/2025", "wmx_2025_4", "WMX", "mx")

rd23_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=54&e=20773&rn=2&rt=M", 2025, 2, 3, "Mason Watt Raceway", "06/08/2025", "wmx_2025_5", "WMX", "mx")

rd31_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20780&rn=1&rt=M", 2025, 3, 1, "Motocross Ste-Julie", "06/29/2025", "wmx_2025_6", "WMX", "mx")

rd32_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20780&rn=2&rt=M", 2025, 3, 2, "Motocross Ste-Julie", "06/29/2025", "wmx_2025_7", "WMX", "mx")

rd41_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20787&rn=1&rt=M", 2025, 4, 1, "Gopher Dunes", "07/06/2025", "wmx_2025_8", "WMX", "mx")

rd42_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20787&rn=2&rt=M", 2025, 4, 2, "Gopher Dunes", "07/06/2025", "wmx_2025_9", "WMX", "mx")

rd51_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20808&rn=1&rt=M", 2025, 5, 1, "Sand Del Lee", "07/13/2025", "wmx_2025_10", "WMX", "mx")

rd52_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20808&rn=2&rt=M", 2025, 5, 2, "Sand Del Lee", "07/13/2025", "wmx_2025_11", "WMX", "mx")

rd61_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20815&rn=1&rt=M", 2025, 6, 1, "Riverglade MX", "07/19/2025", "wmx_2025_11", "WMX", "mx")

rd62_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20815&rn=2&rt=M", 2025, 6, 2, "Riverglade MX", "07/19/2025", "wmx_2025_12", "WMX", "mx")

rd71_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20822&rn=1&rt=M", 2025, 7, 1, "MX Deschambault", "07/26/2025", "wmx_2025_13", "WMX", "mx")

rd72_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20822&rn=2&rt=M", 2025, 7, 2, "MX Deschambault", "07/26/2025", "wmx_2025_14", "WMX", "mx")

rd81_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20829&rn=1&rt=M", 2025, 8, 1, "Walton Raceway", "08/09/2025", "wmx_2025_15", "WMX", "mx")

rd82_2025_wmx <- mx_scraping("https://cmrc.tracksideresults.com/laptimes.asp?s=&c=58&e=20829&rn=2&rt=M", 2025, 8, 2, "Walton Raceway", "08/09/2025", "wmx_2025_16", "WMX", "mx")

all_2025_wmx <- bind_rows(rd11_2025_wmx, rd12_2025_wmx,
                          rd21_2025_wmx, rd22_2025_wmx, rd23_2025_wmx,
                          rd31_2025_wmx, rd32_2025_wmx,
                          rd41_2025_wmx, rd42_2025_wmx,
                          rd51_2025_wmx, rd52_2025_wmx,
                          rd61_2025_wmx, rd62_2025_wmx,
                          rd71_2025_wmx, rd72_2025_wmx,
                          rd81_2025_wmx, rd82_2025_wmx)

all_2025 <- bind_rows(all_2025_450, all_2025_250, all_2025_wmx)
write.csv(all_2025,"all_2025.csv")
```

```{r}
#| warning: false
#| eval: false
all_2025_riders <- all_2025 |>
  distinct(rider)

# List of all riders from the Trackside website
# Type "%" in the search bar
url <- "https://cmrc.tracksideresults.com/search.asp?search=%25&searchType=name&Submit=Find"
webpage <- read_html(url)
riders_table <- webpage |>
  html_nodes("table") |>
  html_table(fill = TRUE)

riders <- riders_table[[6]]
  
riders <- riders |>
  as_tibble() |>
  slice(2:n()) |>
  rename(name = X2, number = X3, location = X4) |>
  select(name, number, location)

write.csv(riders,"all_2025_riders.csv", row.names = FALSE)
```

```{r}
#| warning: false
# In all_2025.csv, replace " " with " " and remove X column (row number). Save as all_2025_cleaned.csv
# In all_2025_riders.csv, I manually added manufacturer, updated_location, round variables. Saved as all_2025_riders_complete.csv

all_2025_cleaned <- read.csv("all_2025_cleaned.csv")
all_2025_riders_complete <- read.csv("all_2025_riders_complete.csv")
```

```{r}
#| warning: false
all_2025_riders_complete <- all_2025_riders_complete |>
  select(-number)

all_2025_complete <- all_2025_cleaned |>
  cross_join(all_2025_riders_complete) |>
  filter(str_detect(rider, name)) |>
  filter(is.na(round.y) | round.x == round.y) |>
  select(finish, name, number, lap, time, behind, place, year, round.x, moto, track, date, race_id, class, type, points, manufacturer, updated_location) |>
  rename(rider = name, round = round.x, location = updated_location) |>
  distinct() |>
  mutate(date = ymd(date))

write.csv(all_2025_complete,"all_2025_complete.csv")
```

New to scraping this year are the bike manufacturers of each rider. 

In future iterations, we want to look at using AI scraping software to help with the process, cutting down time on manual data imputation (currently tedious with ~300 riders).

Is there any way to use Developer Tools in Chrome to work with the tables?

# 3 Analysis

## 3.1 Reapplying 2024 analysis

The goal of this section is to replicate last year's analysis and observe the trends.

### 3.1.1 Demographic summary

Distributions of home locations:

```{r}
#| warning: false
#| echo: false
all_2025_complete |>
  distinct(class, rider, location) |>
  group_by(class, location) |>
  summarize(n = n(),
            .groups = "drop") |>
  ungroup() |>
  arrange(desc(n)) |>
  pivot_wider(names_from = class,
              values_from = n) |>
  filter(! location == "unknown") |>
  mutate(`250` = if_else(is.na(`250`), 0, `250`),
         WMX = if_else(is.na(WMX), 0, WMX),
         `450` = if_else(is.na(`450`), 0, `450`)) |>
  select(location, `450`, `250`, WMX) |>
  gt() |>
  cols_label(location = "Location",
             `450` = "450",
             `250` = "250",
             WMX = "WMX")
```

2025 saw more riders from Ontario (59 vs 46 in 2024). Alberta also experienced a big jump from 22 to 29 and ranking second this year. Locations with fewer riders includes British Columbia (15 vs 26 in 2024).

### 3.1.2 Passing probabilities

<u>450 class</u>:
```{r}
#| warning: false
#| echo: false
all_2025_450_add <- all_2025_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_450_add2 <- all_2025_450_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_450_add <- all_2025_450_add |>
  left_join(all_2025_450_add2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrix450_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_450_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_450_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix450_1[i, j] <- matrix_value
  }
}

rownames(matrix450_1) <- cats
colnames(matrix450_1) <- cats
print(round(matrix450_1, 2))
```

```{r}
#| warning: false
#| echo: false
all_2025_450_placecat <- all_2025_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, 
                            "21-25","26-30"))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", "26-30")))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_450_placecat2 <- all_2025_450_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_450_placecat <- all_2025_450_placecat |>
  left_join(all_2025_450_placecat2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrix450_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_450_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_450_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix450_2[i, j] <- matrix_value
  }
}

rownames(matrix450_2) <- cats2
colnames(matrix450_2) <- cats2

tm_450 <- as.data.frame(as.table(matrix450_2))
ggplot(tm_450, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - 450 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

Compared to 2024: Passing probabilities in the top 5 maintained. However, lower probabilities of staying in the same group from 6th onwards indicating that more passing occured.

<u>250 class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_250_add <- all_2025_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_250_add2 <- all_2025_250_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_250_add <- all_2025_250_add |>
  left_join(all_2025_250_add2) |>
  mutate(prob = n/total)

rows = 40
cols = 40
cats <- 1:40

matrix250_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_250_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_250_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix250_1[i, j] <- matrix_value
  }
}

rownames(matrix250_1) <- cats
colnames(matrix250_1) <- cats
print(round(matrix250_1, 2))
```

```{r}
#| warning: false
#| echo: false
all_2025_250_placecat <- all_2025_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30",
                     ifelse(place >= 31 & place <= 35,
                            "31-35", "36-40"))))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                          ifelse(place_next >= 26 & place_next <= 30, "26-30",
                          ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_250_placecat2 <- all_2025_250_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_250_placecat <- all_2025_250_placecat |>
  left_join(all_2025_250_placecat2) |>
  mutate(prob = n/total)

rows2 = 8
cols2 = 8
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
           "31-35", "36-40")

matrix250_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_250_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_250_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix250_2[i, j] <- matrix_value
  }
}

rownames(matrix250_2) <- cats2
colnames(matrix250_2) <- cats2

tm_250 <- as.data.frame(as.table(matrix250_2))
ggplot(tm_250, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - 250 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

Compared to 2024: This season did not match up to the parity exhibited last year. Showed a similar trend as the 450 class. Lack of passing amongst leaders but more towards the back half of the pack.

<u>WMX class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_wmx_add <- all_2025_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2025_wmx_add2 <- all_2025_wmx_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_wmx_add <- all_2025_wmx_add |>
  left_join(all_2025_wmx_add2) |>
  mutate(prob = n/total)

rows = 32
cols = 32
cats <- 1:32

matrixwmx_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2025_wmx_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_wmx_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrixwmx_1[i, j] <- matrix_value
  }
}

rownames(matrixwmx_1) <- cats
colnames(matrixwmx_1) <- cats
print(round(matrixwmx_1, 2))
```

```{r}
#| warning: false
all_2025_wmx_placecat <- all_2025_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 25 & place <= 30, "26-30",
                            "31-32")))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25",
                          ifelse(place_next >= 26 & place_next <= 30, "26-30","31-32"))))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_wmx_placecat2 <- all_2025_wmx_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_wmx_placecat <- all_2025_wmx_placecat |>
  left_join(all_2025_wmx_placecat2) |>
  mutate(prob = n/total)

rows2 = 7
cols2 = 7
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30", "31-32")

matrixwmx_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_wmx_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_wmx_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrixwmx_2[i, j] <- matrix_value
  }
}

rownames(matrixwmx_2) <- cats2
colnames(matrixwmx_2) <- cats2

tm_wmx <- as.data.frame(as.table(matrixwmx_2))
ggplot(tm_wmx, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2025 Season - WMX class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

Compared to 2024: Last season saw no passes for the lead in the WMX series. So it's easy to see more parity this year as the probability of staying in the lead is 95%, but we do observe more passing throughout the field.

At a track level:
``` {r}
#| warning: false
#| echo: false
pass_prob_moto_fn <- function(input_class, input_track) {
  placecat <- all_2025_complete |>
    filter(class == input_class & track == input_track) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35,
                                                                        "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
    rows = distinct_cat
    cols = distinct_cat
    if (distinct_cat == 4) {
      cats <- c("1-5", "6-10", "11-15", "16-20")
    } else if (distinct_cat == 5) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
    } else if (distinct_cat == 6) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
    } else if (distinct_cat == 7) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35")
    } else if (distinct_cat == 8) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35", "36-40")
    }
    
    pass_matrix <- matrix(0, nrow = rows, ncol = cols)
    for (i in 1:rows) {
      for (j in 1:cols) {
        matrix_row <- placecat |>
          filter(place_cat == cats[i] & place_next_cat == cats[j])
        if (nrow(matrix_row) == 0) {
          matrix_value <- 0
        } else {
          matrix_value <- placecat |> 
            filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
        }
        pass_matrix[i, j] <- matrix_value
      }
    }
    
    rownames(pass_matrix) <- cats
    colnames(pass_matrix) <- cats

  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  p <- ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32", guide = "none") +
    geom_text(aes(label = round(Freq,2)), size = 4) +
    labs(title = "Passing Probabilities",
         subtitle = input_track,
         x = "From Place", 
         y = "To Place", fill = "Probability")
  
  return(p)
}

tracks1 <- all_2025_complete |>
  filter(round <= 4) |>
  distinct(track) |>
  pull(track)

tracks2 <- all_2025_complete |>
  filter(round > 4) |>
  distinct(track) |>
  pull(track)

plots1 <- map(
  tracks1,
  ~ pass_prob_moto_fn(
      input_class = "450",
      input_track = .x
    )
)

plots2 <- map(
  tracks2,
  ~ pass_prob_moto_fn(
      input_class = "450",
      input_track = .x
    )
)

library(patchwork)
wrap_plots(plots1, ncol = 2)
wrap_plots(plots2, ncol = 2)
```

<u>Track insights for the 450 class</u>:
-   Most passing: Mason Watt Raceway
-   Least passing: Motocross Ste-Julie, Sand Del Lee

At an individual rider level (Nicoletti, Ward, Pettis):
```{r}
#| warning: false
#| echo: false
pass_prob_rider_fn <- function(input_rider) {
  placecat <- all_2025_complete |>
    filter(rider == input_rider) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35, "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
  rows = distinct_cat
  cols = distinct_cat
  if (distinct_cat == 2) {
    cats <- c("1-5", "6-10")
  } else if (distinct_cat == 3) {
    cats <- c("1-5", "6-10", "11-15")
  } else if (distinct_cat == 4) {
    cats <- c("1-5", "6-10", "11-15", "16-20")
  } else if (distinct_cat == 5) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
  } else if (distinct_cat == 6) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
  } else if (distinct_cat == 7) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35")
  } else if (distinct_cat == 8) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35", "36-40")
  }
  
  pass_matrix <- matrix(0, nrow = rows, ncol = cols)
  for (i in 1:rows) {
    for (j in 1:cols) {
      matrix_row <- placecat |>
        filter(place_cat == cats[i] & place_next_cat == cats[j])
      if (nrow(matrix_row) == 0) {
        matrix_value <- 0
      } else {
        matrix_value <- placecat |> 
          filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
      }
      pass_matrix[i, j] <- matrix_value
    }
  }
  
  rownames(pass_matrix) <- cats
  colnames(pass_matrix) <- cats
  
  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  p <- ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32", guide = "none") +
    geom_text(aes(label = round(Freq,2)), size = 3) +
    labs(title = "Passing Probabilities",
         subtitle = input_rider, 
         x = "From Place", 
         y = "To Place", fill = "Probability")
  
  return(p)
}

pp_riders <- c("PHIL NICOLETTI", "TANNER WARD", "JESS PETTIS")
plots1 <- map(
  pp_riders,
  ~ pass_prob_rider_fn(
      input_rider = .x
    )
)
wrap_plots(plots1, ncol = 3)
```

In the 2024 report, we looked at 450 rider, Tanner Ward. He is a interesting rider to explore since he finished a close 2nd in points this year and winning 1st overall at a couple events after last season where he finished 5th and beat 5th overall in any event.

In his breakdown last year, he had high probabilities of maintaining in the top 10 and then his probability of going from 11-15 into the top 10 was 29%. This year, he had high probabilities of maintaining in the top 5 and then his probability of going from 6-10 into the top 5 was 11%. It shows that for the 2025 season, he was likely having better starts and when he needed to move up through the field, it was only from 6-10 and not 11-15.

Probability of staying in top 5 throughout moto:
```{r}
#| warning: false
#| echo: false
all_2025_placecat_laps <- all_2025_complete |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place_next) |>
  select(place, place_next, lap, class) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30",
                     ifelse(place >= 31 & place <= 35, "31-35",
                            "36-40"))))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                          ifelse(place_next >= 26 & place_next <= 30, "26-30",
                          ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
  select(place_cat, place_next_cat, lap, class) |>
  group_by(place_cat, place_next_cat, lap, class) |>
  summarise(n = n()) |>
  ungroup()

all_2025_placecat_laps2 <- all_2025_placecat_laps |>
  group_by(class, lap, place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_placecat_laps3 <- all_2025_placecat_laps |>
  left_join(all_2025_placecat_laps2) |>
  mutate(prob = n/total) |>
  filter(place_cat == "1-5" & place_next_cat == "1-5")

all_2025_placecat_laps3 |>
  ggplot(aes(x = lap, y = prob, color = class)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Lap", y = "Probability of Retaining 1-5", color = "Class")
```

Compared to 2024: Similar peak and dip happening around the lap 5 mark. After being the top line last season, WMX appears to be more chaotic and indicates a bump in parity.

Start-finish probabilities:

<u>450 class</u>:
```{r}
#| warning: false
#| echo: false
all_2025_450_start_finish_cat <- all_2025_complete |>
  filter(class == "450" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                           ifelse(place >= 6 & place <= 10, "6-10",
                           ifelse(place >= 11 & place <= 15, "11-15",
                           ifelse(place >= 16 & place <= 20, "16-20",
                           ifelse(place >= 21 & place <= 25, 
                                  "21-25","26-30"))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25", 
                             "26-30")))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_450_start_finish_cat_2 <- all_2025_450_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_450_start_finish_cat_3 <- all_2025_450_start_finish_cat |>
  left_join(all_2025_450_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrix450_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_450_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_450_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrix450_4[i, j] <- matrix_value
  }
}

rownames(matrix450_4) <- cats2
colnames(matrix450_4) <- cats2

sf_450 <- as.data.frame(as.table(matrix450_4))
ggplot(sf_450, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "450", x = "Start ", y = "Finish",
       fill = "Probability")
```

Compared to 2024: Big difference in the 6-10 slot. Higher probabilities of riders falling off down the field from that position.

<u>250 class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_250_start_finish_cat <- all_2025_complete |>
  filter(class == "250" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                           ifelse(place >= 6 & place <= 10, "6-10",
                           ifelse(place >= 11 & place <= 15, "11-15",
                           ifelse(place >= 16 & place <= 20, "16-20",
                           ifelse(place >= 21 & place <= 25, "21-25",
                           ifelse(place >= 26 & place <= 30, "26-30",
                           ifelse(place >= 31 & place <= 35, "31-35", 
                                  "36-40"))))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25", 
                      ifelse(finish >= 26 & finish <= 30, "26-30",
                      ifelse(finish >= 31 & finish <= 35, "31-35", 
                             "36-40")))))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_250_start_finish_cat_2 <- all_2025_250_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_250_start_finish_cat_3 <- all_2025_250_start_finish_cat |>
  left_join(all_2025_250_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 8
cols2 = 8
cats2 <- c("1-5", "6-10", "11-15", "16-20", 
           "21-25", "26-30", "31-35", "36-40")

matrix250_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_250_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_250_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrix250_4[i, j] <- matrix_value
  }
}

rownames(matrix250_4) <- cats2
colnames(matrix250_4) <- cats2

sf_250 <- as.data.frame(as.table(matrix250_4))
ggplot(sf_250, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "250", x = "Start ", y = "Finish",
       fill = "Probability")
```

Compared to 2024: Less parity amongst leading riders. But more movement in the middle of the pack.

<u>WMX class</u>:

```{r}
#| warning: false
#| echo: false
all_2025_wmx_start_finish_cat <- all_2025_complete |>
  filter(class == "WMX" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30", "31-32")))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25",
                      ifelse(finish >= 26 & finish <= 30, "26-30",
                             "31-32"))))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2025_wmx_start_finish_cat_2 <- all_2025_wmx_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2025_wmx_start_finish_cat_3 <- all_2025_wmx_start_finish_cat |>
  left_join(all_2025_wmx_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 7
cols2 = 7
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30", "31-32")

matrixwmx_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2025_wmx_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2025_wmx_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrixwmx_4[i, j] <- matrix_value
  }
}

rownames(matrixwmx_4) <- cats2
colnames(matrixwmx_4) <- cats2

sf_wmx <- as.data.frame(as.table(matrixwmx_4))
ggplot(sf_wmx, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "WMX", x = "Start ", y = "Finish",
       fill = "Probability")
```

Compared to 2024: Greater parity but not as much as expected. Mostly just 1st place that saw more action this season.

Fatigue effects:

<u>450 class</u>:

We filter out lap times over 200 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "450" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2025_complete |>
  filter(class == "450" & time < 200 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "450" & time < 200 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "450 - Average Lap Times")
```

Compared to 2024: Similar trends overall. The 450 class is getting slower over the course of a moto. Interesting how riders were getting faster at both Wild Rose and Prairie Hill. Might have to do with being the first round of the season or track difficulty.

<u>250 class</u>:

We filter out lap times over 250 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "250" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2025_complete |>
  filter(class == "250" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "250" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "250 - Average Lap Times")
```

Compared to 2024: Similar patterns as last season. Fatigue factor was slightly stronger.

<u>WMX class</u>:

We filter out lap times over 300 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "WMX" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2025_complete |>
  filter(class == "WMX" & time < 300 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2025_complete |>
  filter(class == "WMX" & time < 300 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "WMX - Average Lap Times")
```

Compared to 2024: Same trend as last year where for most motos, lap times for riders were likely to decrease in time. Unlike the men's classes, this is an indication that women riders are not fatiguing since they are getting faster.

Another way for us to assess fatigue is to compare lap times during the first motos with lap times of the second moto. To make the analysis less clutered, we look at the overall class level. Because this involves motos at different tracks, looking at raw lap times doesn't make sense. So we define and use decay percentage.

$$Decay \ pct = \left( \frac{\text{time} - \text{fastest moto lap time}}{\text{fastest moto lap time}} \right) \times 100$$

For this analysis, we'll look at the top 15 riders in the each moto. To account for outliers, we only select lap times less than 300 seconds and remove the first and last lap. 

```{r}
#| warning: false
all_2025_complete |>
  filter(time < 300 & lap > 1 & finish <= 15) |>
  group_by(rider, race_id) |>
  mutate(fastest_lap_time = min(time, na.rm = TRUE),
         decay_pct = ((time - fastest_lap_time) / fastest_lap_time) * 100) |>
  ungroup() |>
  group_by(lap, class) |>
  summarize(avg_decay_pct = mean(decay_pct)) |>
  ungroup() |>
  group_by(class) |>
  arrange(lap) |>
  slice(1:(n() - 1)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_decay_pct, 
             color = class)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  labs(x = "Lap", y = "Avg Decay Pct", color = "Class",
       title = "Avg Decay Pct by Class")
```

This shows that after the lap 5 mark, riders in the 450 and 250 class really start to decay. Riders in the WMX class start to decay sooner on average but the decay is much more volitile. 

```{r}
#| warning: false
decay_pcts <- all_2025_complete |>
  filter(time < 300 & lap > 1 & finish <= 15) |>
  group_by(rider, race_id) |>
  mutate(fastest_lap_time = min(time, na.rm = TRUE),
         decay_pct = ((time - fastest_lap_time) / fastest_lap_time) * 100) |>
  ungroup() |>
  group_by(class, moto, lap) |>
  summarize(avg_decay_pct = mean(decay_pct)) |>
  ungroup() |>
  group_by(class, moto) |>
  arrange(lap) |>
  slice(1:(n() - 1)) |>
  ungroup()

decay_pcts |>
  filter(class == "450") |>
  mutate(moto = as.factor(moto)) |>
  ggplot(aes(x = lap, y = avg_decay_pct, 
             color = moto)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  labs(x = "Lap", y = "Avg Decay Pct", color = "Moto",
       title = "450 - Avg Decay Pct by Moto")
```

This shows that for 450 riders, the decay rate is more prominent during the second moto. This is likely because they are less fresh compared to the first moto.

```{r}
#| warning: false
decay_pcts |>
  filter(class == "250") |>
  mutate(moto = as.factor(moto)) |>
  ggplot(aes(x = lap, y = avg_decay_pct, 
             color = moto)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  labs(x = "Lap", y = "Avg Decay Pct", color = "Moto",
       title = "250 - Avg Decay Pct by Moto")
```

This shows pretty much the same thing for 250 riders. Decay rate is stronger for moto 2 but the inflection point happens later on average in moto 2 (around lap 6).

```{r}
#| warning: false
decay_pcts |>
  filter(class == "WMX") |>
  mutate(moto = as.factor(moto)) |>
  ggplot(aes(x = lap, y = avg_decay_pct, 
             color = moto)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  labs(x = "Lap", y = "Avg Decay Pct", color = "Moto",
       title = "WMX - Avg Decay Pct by Moto")
```

No noticeable patterns to observe for the women's series.

## 3.2 Moto quartile positions
The first thing we need to do to determine a rider's position at the 0/25/50/75/100% mark of the moto is finding out what laps these occur on. 

In this first block, we determine a particular rider's positions for a given moto. We'll use Tanner Ward during the final moto of the season at Walton for example.

```{r}
#| warning: false
quartile_positions <- all_2025_complete |>
  group_by(rider, race_id) |>
  mutate(max_lap = max(lap),
         lap_25 = round(max_lap * 0.25),
         lap_50 = round(max_lap * 0.50),
         lap_75 = round(max_lap * 0.75),
         lap_milestone = case_when(
         lap == 1 ~ "Start",
         lap == lap_25 ~ "25%",
         lap == lap_50 ~ "50%",
         lap == lap_75 ~ "75%",
         TRUE ~ NA_character_)) |>
  ungroup() |>
  filter(!is.na(lap_milestone))

quartile_positions_moto_fn <- function(select_rider, select_moto_id) {
  quartile_positions |> 
    select(rider, race_id, round, moto, lap_milestone, place, finish) |>
    pivot_wider(
    names_from = lap_milestone, 
    values_from = place) |> 
    filter(rider == select_rider & race_id == select_moto_id) |>
    select(rider, race_id, round, moto, Start, `25%`, `50%`, `75%`, finish) |>
    rename(Finish = finish)
}
quartile_positions_moto_fn("TANNER WARD", "450_2025_16")
```

In the next block, we calculate a particular rider's season averages for their places at these quartiles.

```{r}
#| warning: false
quartile_positions_avg_fn <- function(select_rider) {
  quartile_positions |> 
    select(rider, race_id, round, moto, lap_milestone, place, finish) |>
    pivot_wider(
    names_from = lap_milestone, 
    values_from = place) |> 
    filter(rider == select_rider) |>
    select(rider, race_id, round, moto, Start, `25%`, `50%`, `75%`, finish) |>
    rename(Finish = finish)
}
quartile_positions_avg_fn("TANNER WARD")
```

Summary of Tanner Ward's season: Decent starts and worked his way to a higher place most of the motos.

## 3.3 Standarized lap times
Need a way to compare across riders and tracks.
## 3.4 Consistency scores
Separate steady from volatile riders. Need to apply lap time standardization
## 3.5 Defensive skill
Ability to hold position under pressure (rider consistently within 2 seconds back).
## 3.6 Inflection laps
Revisiting win probability model from last year
More evaluation to make sure it is accurate
Identify laps where win probability shifts sharply
## 3.7 Traffic effects
Clear air vs battles: Top riders standardized times, split between when in 1st place and trailing someone by 2 seconds.
What about when leaders come up on lap traffic? Blue flag - does it work or are leaders slowed down?
## 3.8 Gender gap
Compare 250 class and WMX class riders since they use the same bike. How would the women do in the men's series (simulation)?
## 3.9 Prize money
Is it a financially viable occupation? Need to reference official series docs.
## 3.10 Entertainment value
Average gap to leader by series
## 3.11 Weather effects
Compare the lap times for this year to last year at the same tracks with the temperature. Consider weather types (like rain) and track changes (not the same setup every year)
## 3.12 Fall prediction
First, creating the reasonable definition for a fall - big delta in lap time and place, big difference between last lap place and finish position. No easy way to confirm.
Then look at the characteristics/conditions around those falls - which riders typically involved, late in the moto?
What would happen in the points race if certain accidents (falls of 2 riders at same time) didn't happen?
## 3.13 Rider profiles
Conducted in Tableau this time. Include link here.
Archetype clusters -> Look at the NHL/NBA player cards (percentiles) and try to replicate.
Include photos to display.
## 3.14 Expected finish model
Linear regression/random forest?
Finish ~ Rider historical pace + lap + current place + time gaps to each place
## 3.15 Pass probability model
P(place-1) ~ Rider historical pace? + lap + current place + time behind next place + time ahead of next place

# 4 Extensions
The big thing I want to do to expand the project in the future is add previous years data to the analysis. Consolidate everything in one dataset. According to the [results](https://triplecrownseries.ca/results/) tab of the Triple Crown Series website, there is data going back to 2018.

# 5 Notes
Hi, I'm Bryanna and I love pivottables. I got passed around as a baby.